Future challenges for Research Software Engineering

### 2025-05-17 future_challenges_for_RSEng.tex
Update future_challenges_for_RSEng.tex
Co-authored-by: Magnus Hagdorn <Magnus.Hagdorn@marsupium.org>

% Article frontmatter
% ===================
\title{Future challenges for Research Software Engineering} % Title of the article
%\short{} % Short title of the article (optional)
\newcommand{\authorRef}[1]{\texorpdfstring{\autref{#1}}{}}
\author{
Florian Goth\authorRef{1},
Leyla Jael Castro\authorRef{2},
Gerasimos Chourdakis\authorRef{3},
Simon Christ\authorRef{4},
Jeremy Cohen\authorRef{5},
Jean-Noël Grad\authorRef{6},
Magnus Hagdorn\authorRef{7},
Toby Hodges\authorRef{8},
Jan Linxweiler\authorRef{9},
Frank Löffler\authorRef{10},
Michele Martone\authorRef{11},
Jan Philipp Thiele\authorRef{9}\textsuperscript{,}\authorRef{12},
Nicholas A. Del Grosso\authorRef{13}
} % Authors and references to addresses
\institute{% Institutes with labels
\autlabel{1} Institut für theoretische Physik 1, University of Würzburg, 97074, Würzburg, Germany\par
\autlabel{2} ZB MED Information Centre for Life Sciences, Cologne, Germany\par
\autlabel{3} Institute for Parallel and Distributed Systems, University of Stuttgart, Stuttgart, Germany\par
\autlabel{4} Leibniz University Hannover, Department of Cell Biology and Biophysics, Computational Biology, Germany\par
\autlabel{5} Imperial College London, London, UK\par
\autlabel{6} Institute for Computational Physics, University of Stuttgart, Germany\par
\autlabel{7} Geschäftsbereich IT, Charité Universitätsmedizin Berlin, Germany\par
\autlabel{8} The Carpentries, USA\par
\autlabel{9} Technische Universität Braunschweig, Germany\par
\autlabel{10} Michael Stifel Center Jena {\&} Friedrich Schiller University Jena, Germany\par
\autlabel{11} Leibniz Supercomputing Centre, Garching, Germany\par
\autlabel{12} Weierstrass Institute, Berlin, Germany;
              Leibniz University Hannover, Institute of Applied Mathematics, Scientific Computing, Hannover, Germany\par
\autlabel{13} Institute for Experimental Epileptology and Cognition Research, Uniklinikum Bonn, Germany
}
\abstract{
It is 2025 and we are, admittedly, still working on establishing and growing Research Software Engineering (RSEng)
as a domain in its own right, with the hope of ensuring that it gets the recognition it deserves.
Nonetheless, as the use of digital tools becomes ever more pervasive in a researcher's day-to-day tasks, there are already people suggesting that RSEng will
cease to exist as a separate field, and the specialist skills that Research Software Engineers(RSEs) provide to support and undertake research will
become a core part of the tooling and skill set of researchers in almost all other domains.
Are we already approaching ``the end of Research Software Engineering''?!
While we believe that it is certainly too early to already herald the demise of RSE as a separate field,
we also recognise the need to acknowledge that the future is in constant flux and
it is worthwhile discussing how RSEng will/could change in response to these upcoming challenges.
To figure out these challenges of the future we held a workshop and structured the resulting discussion around social, technological, and organisational aspects.
} % Abstract of the article
\keywords{Keywords go here.} % Keywords for the article

@@

Research Software Engineering is still a developing field,
with increasing integration into various aspects of research.
In this opinion paper, we try to imagine the state of our field
and community in the mid-term future, focusing on future challenges.
While future projections are common in industry \cite{gartner}, so far there is a projection lacking for the RSEng space.
We present here the results of a workshop discussion at the
deRSE25 conference in Karlsruhe\cite{Goth2025EndRSEng}, without
claiming to represent the entire deRSE community.
Nevertheless, by disseminating the discussed ideas,
we aim to engage the community in creating a future worth living
and advance the conversation about how Research Software Engineering should, or could, develop.\\
It is already difficult to structure the spectrum of RSE tasks and responsibilities today given the spectrum of research software \cite{hasselbring2024}
and it is even more challenging to structure the aspects that will become relevant for the RSE community in the future.
Nevertheless, we decided to consider:
\begin{itemize}
\item The evolution of the term RSE and its definition
\item Growth and development of the RSE community
\item The impact of automation and generative artificial intelligence in RSE work
\item The impact of complexity and usability in research software
\item Ethics and social considerations
\item The transformations expected in academic institutions
\item Impact of alternate career paths for RSEs and other RTPs(\emph{Research Technical Professional}s)
\end{itemize}
We conclude the paper with a summary and with our wishes and recommendations.

>>>>

% Article frontmatter
% ===================
\title{Future challenges for Research Software Engineering} % Title of the article
%\short{} % Short title of the article (optional)
\newcommand{\authorRef}[1]{\texorpdfstring{\autref{#1}}{}}
\author{
Florian Goth\authorRef{1},
Leyla Jael Castro\authorRef{2},
Gerasimos Chourdakis\authorRef{3},
Simon Christ\authorRef{4},
Jeremy Cohen\authorRef{5},
Jean-Noël Grad\authorRef{6},
Magnus Hagdorn\authorRef{7},
Toby Hodges\authorRef{8},
Jan Linxweiler\authorRef{9},
Frank Löffler\authorRef{10},
Michele Martone\authorRef{11},
Jan Philipp Thiele\authorRef{9}\textsuperscript{,}\authorRef{12},
Nicholas A. Del Grosso\authorRef{13}
} % Authors and references to addresses
\institute{% Institutes with labels
\autlabel{1} Institut für theoretische Physik 1, University of Würzburg, 97074, Würzburg, Germany\par
\autlabel{2} ZB MED Information Centre for Life Sciences, Cologne, Germany\par
\autlabel{3} Institute for Parallel and Distributed Systems, University of Stuttgart, Stuttgart, Germany\par
\autlabel{4} Leibniz University Hannover, Department of Cell Biology and Biophysics, Computational Biology, Germany\par
\autlabel{5} Imperial College London, London, UK\par
\autlabel{6} Institute for Computational Physics, University of Stuttgart, Germany\par
\autlabel{7} Geschäftsbereich IT, Charité Universitätsmedizin Berlin, Germany\par
\autlabel{8} The Carpentries, USA\par
\autlabel{9} Technische Universität Braunschweig, Germany\par
\autlabel{10} Michael Stifel Center Jena {\&} Friedrich Schiller University Jena, Germany\par
\autlabel{11} Leibniz Supercomputing Centre, Garching, Germany\par
\autlabel{12} Weierstrass Institute, Berlin, Germany;
              Leibniz University Hannover, Institute of Applied Mathematics, Scientific Computing, Hannover, Germany\par
\autlabel{13} Institute for Experimental Epileptology and Cognition Research, Uniklinikum Bonn, Germany
}
\abstract{
It is 2025 and we are, admittedly, still working on establishing and growing Research Software Engineering (RSEng)
as a domain in its own right, with the hope of ensuring that it gets the recognition it deserves.
Nonetheless, as the use of digital tools becomes ever more pervasive in a researcher's day-to-day tasks, there are already people suggesting that RSEng will
cease to exist as a separate field, and the specialist skills that Research Software Engineers(RSEs) provide to support and undertake research will
become a core part of the tooling and skill set of researchers in almost all other domains.
Are we already approaching ``the end of Research Software Engineering''?!
While we believe that it is certainly too early to already herald the demise of RSE as a separate field,
we also recognise the need to acknowledge that the future is in constant flux and
it is worthwhile discussing how RSEng will/could change in response to these upcoming challenges.
To figure out these challenges of the future we held a workshop and structured the resulting discussion around social, technological, and organisational aspects.
} % Abstract of the article
\keywords{Future RSE, Evolving field, Vision} % Keywords for the article

@@

Research Software Engineering is still a developing field,
with increasing integration into various aspects of research.
In this opinion paper, we try to imagine the state of our field
and community in the mid-term future, focusing on future challenges.
While future projections are common in industry \cite{gartner}, so far there is a projection lacking for the RSEng space.
We present here the results of a workshop discussion at the
deRSE25 conference in Karlsruhe\cite{Goth2025EndRSEng}, without
claiming to represent the entire deRSE community.
Nevertheless, by disseminating the discussed ideas,
we aim to engage the community in creating a future worth living
and advance the conversation about how Research Software Engineering should, or could, develop.\\
It is already difficult to structure the spectrum of RSE tasks and responsibilities today given the spectrum of research software \cite{hasselbring2024}
and it is even more challenging to structure the aspects that will become relevant for the RSE community in the future.
Nevertheless, we decided to consider:
\begin{itemize}
\item The evolution of the term RSE and its definition
\item Growth and development of the RSE community
\item The impact of automation and generative artificial intelligence in RSE work
\item The impact of complexity and usability in research software
\item Ethics and social considerations
\item The transformations expected in academic institutions
\item Impact of alternative career paths for RSEs and other RTPs(\emph{Research Technical Professional}s)
\end{itemize}
We conclude the paper with a summary and with our wishes and recommendations.
